Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-13 paper/ResponseToReviews.tex

What would first do is add up the transcript counts for well known markers of actively proliferating cells (CCNB2 is my favorite, but I you could sum a bunch of cyclins and CDKs) to get a ?proliferation index? for each cell. Then look at the distribution of this - I would expect it to be lognormal for a population of cells that are robustly growing. For cells that are mostly quiescent (e.g. a confluent epithelial plate), I would guess the transcript levels for the cells would be low, maybe with a long tail to the right. Then you would add the proliferation index values as a new column in the pData table in Monocle. Then you can use the index as a variable in the model formula for differentialGeneTest. This will give you genes that positively correlate with the proliferation index.
The trouble is establishing what values of the index really correspond to actively cycling cells. Ideally, you?d have a bimodal distribution and you?d be able to set some clear threshold, but in practice this doesn?t usually happen. In general, this is pretty tricky to do without built-in controls (e.g. sets of cells that are sorted on DNA content, arrested with some drug, or what have you).
So I am generally incline to take these sort of analyses with a huge grain of salt.
}

- In assessing to what extent lack of RNP expression accounts for the viral mRNA expression variability (page 6, paragraph 2), we think it is important to take into account potential extracellular contamination. Specifically, it would be more convincing if the analysis omitted cells with mixed wildtype/synonymous clones.

{\color{black}
This is a good suggestion.
We have re-performed the analyses omitting the co-infected cells with mixed viral barcodes.
These new analyses are in Figure 5-Figure supplement 3 and Figure 5-Figure supplement 5.
The results remain essentially unchanged from the ones analyzing the full set of cells that are shown in main panels of Figure 5A,B.
}

### 2018-01-13 paper/paper.tex
repeat Figure 5 analyses (RNP effect on viral expression) excluding co-infected cells

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
However, an inherent limitation of current established single-cell mRNA sequencing techniques is that they only sequence the 3' end of the transcript~\citep{zheng2017massively,macosko2015highly,Klein:2015ki,cao2017comprehensive}.
Since the alternative spliceoforms M1 / M2 and NS1 / NEP share the same 3' ends, we cannot distinguish them and therefore will refer simply to the combined counts of transcripts from each of these alternatively spliced segments as the M and NS genes.

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).

>>>>

\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
Plots for individual samples are in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}.
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
The trends are similar if we look only at the 10-hour sample (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}) or exclude known co-infected cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdenbyflugene_nocoinfection}).
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (A), but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburden_nocoinfection}]{
The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, excluding cells that were annotated as coinfected based on the presence of both viral barcodes.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_no_coinfect.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figsupp[Like panel (B) but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburdenbyflugene_nocoinfection}]{
All findings in Figure~\ref{fig:fluburdenbyflugene}B remain unchanged if we exclude cells called as coinfected based on the presence of mixed viral barcodes.
}{\includegraphics[width=\linewidth]{figures/p_no_coinfect_flu_burden_flu_gene_test}}
\figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}). 
Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
However, an inherent limitation of current established single-cell mRNA sequencing techniques is that they only sequence the 3' end of the transcript~\citep{zheng2017massively,macosko2015highly,Klein:2015ki,cao2017comprehensive}.
Since the alternative spliceoforms M1 / M2 and NS1 / NEP share the same 3' ends, we cannot distinguish them and therefore will refer simply to the combined counts of transcripts from each of these alternatively spliced segments as the M and NS genes.

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
